Component response rate variation drives stability in large complex systems

### 2018-05-21 ms.Rmd
Add website
More writing

arxiv comand: tar cvfz filename.tgz filename.tex filename_files
--->

@@

Higher stability in systems with variation in $\gamma$ can be observed by sampling $\gamma_{i}$ values from various distributions. I now focus on a uniform distribution where $\gamma \sim \mathcal{U}(0, 2)$ (see Supporting Information for other distributions, which give similar results). As with the case of $\gamma = \{1.95, 0.5\}$ (Fig. 1b), $E[\gamma] = 1$ when $\gamma \sim \mathcal{U}(0, 2)$, allowing comparison of $M$ before and after variation in component response rate. Figure 2 shows a comparison of eigenvalue distributions given $S = 1000$, $C = 1$, and $\sigma = 0.4$. As expected [@Tao2010], when $\gamma = 1$, eigenvalues are distributed uniformly in a circle centred at ($-1, 0$) with a radius of $\sigma\sqrt{SC} =$ `r round(sigma*sqrt(S*C), digits = 3)`. Uniform variation in $\gamma$ leads to a non-uniform distribution of eigenvalues, some of which are clustered tightly around the centre of the distribution, but others of which are spread outside the former radius of `r round(sigma*sqrt(S*C), digits = 3)` (red circle Fig 2b). This larger radius occurs because the addition of $Var(\gamma)$ increases the realised $\sigma$ of $M$. The clustering and spreading of eigenvalues introduced by $Var(\gamma)$ can destabilise previously stable systems or stabilise systems that are otherwise unstable. But where systems are otherwise too complex to be stable given $\gamma = 1$, the effect of $Var(\gamma)$ can often lead to stability above May's[@May1972; @Allesina2012] threshold of $\sigma\sqrt{SC} > 1$.

@@

To investigate the effect of $Var(\gamma)$ on system stability, I simulated random $M$ matrices at $\sigma = 0.4$ and $C = 1$ across $S$ ranging from $2-32$ (see Supporting Information for different values of $\sigma$ and $C$). One million $M$ were simulated for each $S$, and the stability of $M$ was assessed given $\gamma = 1$ versus $\gamma \sim \mathcal{U}(0, 2)$ (note that under these conditions, $\sigma\sqrt{SC} = 1$ given $S = 25$ when $\gamma = 1$). I found that the number of stable random systems was consistently higher given $Var(\gamma)$ than when $\gamma = 1$ (Fig. 3), and that the difference between the probabilities of observing a stable system increased with an increase in $S$; i.e., the potential for $Var(\gamma)$ to drive stability increased with system complexity. For the highest values of $S$, nearly all systems that were stable given $Var(\gamma)$ would not have been stable given $\gamma = 1$, and the maximum observed $S$ for which a system was stable was $31$ given $Var(\gamma)$ versus $27$ given $\gamma = 1$ (see Supporting Information for full results). This suggests that the stability of large systems might be dependent upon variation in the response rate of their individual components, meaning that factors such as generation time (in ecological networks), transaction speed (in economic networks), or communication speed (in social networks) needs to be considered when investigating the stability of complex systems.

Some care is needed when interpreting these results. First, I emphasise that $Var(\gamma)$ is not stabilising per se; that is, adding variation in $\gamma$ to a particular system $M$ does not necessarily increase the probability that the system will be stable (see Supporting Information). Rather, systems that are observed to be stable are more likely to vary in $\gamma$, and for this $Var(\gamma)$ to be critical to their stability. This is caused by the shift in the distribution of eigenvalues that occurs by introducing $Var(\gamma)$ (Fig. 1b, 2b), which can sometimes result in all $\Re(\lambda) < 0$ but might also increase $\Re(\lambda)$ values. <!--- Mention specific ecological systems here? --->

@@

6. A paragraph explaining the results of the random matrices, with brief mention of other types of systems (e.g., predator-prey, mutualist, competitor) presented in supporting information. The obvious increase in stability with different variation of $\gamma_{i}$ will be presented, along with an intuitive explanation of why this is the case (still waiting on simulations -- it would be great if correlation between positive or negative processes and response times led to increased stability, as appears to be the case. I could then argue that $Var(\gamma_{i})$ has a minimum such correlation of zero, and can only increase in stabilisation by having an absolute value of correlation being greater than zero). At least two [key figures](#fig_ideas) will be included, likely with four panels (1. without variation, 2. with unstructured variation, and 3. and 4. with structured variation).

>>>>

arxiv comand: tar cvfz filename.tgz filename.tex filename_files

https://gist.github.com/JJ/ef9d3d8f1142df064bd5
--->

@@

Higher stability in systems with variation in $\gamma$ can be observed by sampling $\gamma_{i}$ values from various distributions. I now focus on a uniform distribution where $\gamma \sim \mathcal{U}(0, 2)$ (see Supplementary Information for other distributions, which give similar results). As with the case of $\gamma = \{1.95, 0.5\}$ (Fig. 1b), $E[\gamma] = 1$ when $\gamma \sim \mathcal{U}(0, 2)$, allowing comparison of $M$ before and after variation in component response rate. Figure 2 shows a comparison of eigenvalue distributions given $S = 1000$, $C = 1$, and $\sigma = 0.4$. As expected [@Tao2010], when $\gamma = 1$, eigenvalues are distributed uniformly in a circle centred at ($-1, 0$) with a radius of $\sigma\sqrt{SC} =$ `r round(sigma*sqrt(S*C), digits = 3)`. Uniform variation in $\gamma$ leads to a non-uniform distribution of eigenvalues, some of which are clustered tightly around the centre of the distribution, but others of which are spread outside the former radius of `r round(sigma*sqrt(S*C), digits = 3)` (red circle Fig 2b). This larger radius occurs because the addition of $Var(\gamma)$ increases the realised $\sigma$ of $M$. The clustering and spreading of eigenvalues introduced by $Var(\gamma)$ can destabilise previously stable systems or stabilise systems that are otherwise unstable. But where systems are otherwise too complex to be stable given $\gamma = 1$, the effect of $Var(\gamma)$ can often lead to stability above May's[@May1972; @Allesina2012] threshold of $\sigma\sqrt{SC} > 1$.

@@

To investigate the effect of $Var(\gamma)$ on system stability, I simulated random $M$ matrices at $\sigma = 0.4$ and $C = 1$ across $S$ ranging from $2-32$ (see Supplementary Information for different values of $\sigma$ and $C$). One million $M$ were simulated for each $S$, and the stability of $M$ was assessed given $\gamma = 1$ versus $\gamma \sim \mathcal{U}(0, 2)$ (note that under these conditions, $\sigma\sqrt{SC} = 1$ given $S = 25$ when $\gamma = 1$). I found that the number of stable random systems was consistently higher given $Var(\gamma)$ than when $\gamma = 1$ (Fig. 3), and that the difference between the probabilities of observing a stable system increased with an increase in $S$; i.e., the potential for $Var(\gamma)$ to drive stability increased with system complexity. For the highest values of $S$, nearly all systems that were stable given $Var(\gamma)$ would not have been stable given $\gamma = 1$, and the maximum observed $S$ for which a system was stable was $31$ given $Var(\gamma)$ versus $27$ given $\gamma = 1$ (see Supplementary Information for full results). This suggests that the stability of large systems might be dependent upon variation in the response rate of their individual components, meaning that factors such as generation time (in ecological networks), transaction speed (in economic networks), or communication speed (in social networks) needs to be considered when investigating the stability of complex systems.

Some care is needed when interpreting these results. First, I emphasise that $Var(\gamma)$ is not stabilising per se; that is, adding variation in $\gamma$ to a particular system $M$ does not necessarily increase the probability that the system will be stable (see Supplementary Information). Rather, systems that are observed to be stable are more likely to vary in $\gamma$, and for this $Var(\gamma)$ to be critical to their stability. This is caused by the shift in the distribution of eigenvalues that occurs by introducing $Var(\gamma)$ (Fig. 1b, 2b), which can sometimes result in all $\Re(\lambda) < 0$ but might also increase $\Re(\lambda)$ values. To further investigate the potential of $Var(\gamma)$ to be stabilising, I used a simple genetic algorithm because the space of possible $\gamma$ values was too large to searh exhaustively[@Hamblin2013] (see Supplementary Information). For each of 10000 random $M$, the genetic algorithm initialised 1000 different sets of $\gamma \sim \mathcal{U}(0, 2)$ values of size $S$. Eigenanalysis of each set of $\gamma$ values was performed, and the 20 sets with the lowest $\max\left(\Re(\lambda)\right)$ each produced 50 offspring with subsequent mutation and crossover between the resulting new population of 1000 $\gamma$ sets. The genetic algorithm terminated if a stable $M$ was found, 20 generations occurred, or a converge criteria of minimum fitness increase was satisfied. Across $S = \{2, 3, ..., 49, 50\}$, sets of $\gamma$ values were frequently found result in stable systems, suggesting that varying component response time might by itself be sufficient to stabilise complex systems. <!--- Fix these last few sentences --->

I have focused broadly on random complex systems, but

My results show that complex systems are more likely to be stable when the response rates of system components vary. These results are broadly applicable to complex biological and social networks.

@@

6. A paragraph explaining the results of the random matrices, with brief mention of other types of systems (e.g., predator-prey, mutualist, competitor) presented in Supplementary Information. The obvious increase in stability with different variation of $\gamma_{i}$ will be presented, along with an intuitive explanation of why this is the case (still waiting on simulations -- it would be great if correlation between positive or negative processes and response times led to increased stability, as appears to be the case. I could then argue that $Var(\gamma_{i})$ has a minimum such correlation of zero, and can only increase in stabilisation by having an absolute value of correlation being greater than zero). At least two [key figures](#fig_ideas) will be included, likely with four panels (1. without variation, 2. with unstructured variation, and 3. and 4. with structured variation).
